Predicting the effects of character displacement on food-web dynamics

### 2019-02-03 manuscript_ECD_model.Rmd
many changes plus working on Discussion content.

---
title: Predicting the effects of character displacement on food-web dynamics
author: 
- Matthew A. Barbour$^{1,2,\ast}$
# potentially use for 'date' for affiliations, as recommended by https://github.com/rstudio/bookdown/issues/348 for cross-format compatibility, with | to create new sections.
#date: | 
#    | 1. Affiliation
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    #citation_package: natbib
    template: Rmd_AmNat_MS_template.tex
    includes:  
      in_header: Am-Nat-preamble-latex.tex
bibliography: references_ECD_supp_mat.bib
biblio-style: amnat.bst
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
---

@@

\noindent 1. University of British Columbia, Department of Zoology, Vancouver, BC V6T 1Z4, Canada;

\noindent 2. University of Zurich, Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich, Switzerland;

@@

*Manuscript elements*: Figure 1, figure 2, figure 3, figure 4. All figures should be printed in color.

@@

Although explicit models of resources gave key insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. This is likely because character displacement has been primarily studied through the lens of competition and coexistence theory [Lawlor1976; Germain2018]. For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. On the other hand, specialization often increases a consumer’s ability to capture a specific resource. A well known consequence of increased foraging efficiency is resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011]. NEED A LAST SENTENCE TO WRAP IT UP, HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY. PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP PREDATOR (I.E. NO CONSUMERS...LOOK AT OLD TEXT FOR WORDING SUGGESTIONS.)

@@

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972; @Lawlor1976); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

@@

To examine how ecological character displacement affects consumer-resource interactions and food-web dynamics, I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the scenario where consumer feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$, then this model becomes MacArthur's model of resource competition (@MacArthur1972). The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail (@Lawlor1976; @Abrams1986), with the general result being divergent character displacement. I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, defined as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer. I focus here on the ecological consequences of this adaptive divergence.

@@

I examine the effects of this inevitable divergence in attack rates on food-web dynamics by (i) deriving analytical expressions for the relationship between attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement. <!--, rather than any dependence on inherent asymmetries between consumers or resources.-->

@@

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:Tradeoff}) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:Tradeoff}). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:Tradeoff}). Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0).

@@

```{r Tradeoff, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Tradeoff}Tradeoff forms for consumer attack rates."}
tradeoff_plot <- bind_rows(concaveDN_df, linear_df, concaveUP_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)

@@

My simulations show that the shape of the attack-rate tradeoff qualitatively affects the relationship between character displacement and resource density (Fig. \ref{fig:MacArthur_Resources}). For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:MacArthur_Resources}A) and character displacement has no effect on resource densities (Fig. \ref{fig:MacArthur_Resources}B). If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:MacArthur_Resources}B). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Resources}A). When the tradeoff is concave up (orange lines), character displacement suppresses resource densities due to the increase in total attack rates (Fig. \ref{fig:MacArthur_Resources}A,B). Although the equation I derived for resource density was for the scenario where both consumers and both resources are present, it accurately predicts the density of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Resources}B). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Resources}A), resulting in equivalent resource densities.

The effect of character displacement on resource density closely matches its impact on food-web stability (Fig. \ref{fig:Stability}). For example, when character displacement results in resource suppression (orange), there is a corresponding decrease in stability. Similarly, if character displacement does not influence resource densities (blue), then there is no corresponding effect on stability. Curiously, this match between impact on resources and stability breakdowns with the concave down tradeoff (green). <!--This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:Stability}). The effect of this eco-evolutionary feedback can be seen by comparing initial (small points) with evolved strategies (large points). For example, when the tradeoff is concave up (orange points), the three-species system (small triangle) is initially more stable than the four-species system (small circle), but this relationship reverses by the time consumers reach their evolutionary stable strategy.-->

```{r MacArthur_Resources, include=FALSE, echo=FALSE, fig.cap="\\label{fig:MacArthur_Resources}Effect of character displacement on total attack rates (A) and resource densities (B) when consumers can forage for both resources simultaneously."}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=factor(n), shape=Competitor), size=3, color="black") + #, show.legend=FALSE) + size=5,
  #geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=a11+a12, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  ylab(expression(a[ii]+a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability, include=FALSE, echo=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="black", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=5, color="black", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability ("-lambda)) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r MacArthur Model, echo=FALSE, fig.cap="New MacArthur model figure"}
plot_macarthur_model <- plot_grid(
  tradeoff_plot + theme(legend.position = "none"), 
  Mac_attack  + theme(legend.position = "none"), 
  Mac_resources  + theme(legend.position = "none"),
  Stability_Mac  + theme(legend.position = "none"), ncol=2, align='hv')

@@

```{r MacArthur_Plot, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:MacArthur_Plot}Fill in."}
knitr::include_graphics("Figure_ECD_MacArthur.pdf")
```

@@

The only character displacement model that I am aware of that modeled resources in different habitats was one examined by @Lawlor1976. This model takes the same form as the previous model, except now the consumer's feeding rate takes the form:

@@

where $w_{ji}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $j$ is found (i.e. habitat preference). Note that since $w_{ji}$ is a proportion that $w_{jj}=1-w_{ji}$. As with attack rates, I assume that consumer habitat preferences are mirror images of each other ($w=w_{11}=w_{22}$). In @Lawlor1976's detailed evolutionary analysis of this model, they again always observe divergent character displacement.

@@

To examine the ecological consequences of this model, I again derive the equilibrium solution when both consumers and both resources are present. Again, resource densities are equivalent at this equilibrium ($\hat R = \hat R_1 = \hat R_2$ when both consumers and resources present), but are now determined by the following equation (derivation in supplementary material):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:LS_Resources}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Resources}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Resources}B). In contrast, resource densities reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Resources}B), because character displacement tends to reach a constraint of complete specialization. It is worth noting that in this foraging scenario, resource densities are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Resources}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Resources}B). <!-- I will revisit this subtle point when I examine the effects of character displacement on food-web stability. -->

>>>>

---
title: Predicting the effects of character displacement on food-web dynamics
author: 
- Matthew A. Barbour$^{1,2,\ast}$
# potentially use for 'date' for affiliations, as recommended by https://github.com/rstudio/bookdown/issues/348 for cross-format compatibility, with | to create new sections.
#date: | 
#    | 1. Affiliation
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    #citation_package: natbib # why isn't this working?
    template: Rmd_AmNat_MS_template.tex
    includes:  
      in_header: Am-Nat-preamble-latex.tex
bibliography: references.bib
biblio-style: amnat.bst
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
citecolor: black
---

@@

\noindent 1. University of Zurich, Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich, Switzerland;

\noindent 2. University of British Columbia, Department of Zoology, Vancouver, BC V6T 1Z4, Canada;

@@

*Manuscript elements*: Figure 1, figure 2, and figure 3. All figures should be printed in color.

@@

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [Lawlor1976; Germain2018; Bassar; McPeek]. For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. On the other hand, specialization often increases a consumer’s ability to capture a specific resource. A well known consequence of increased foraging efficiency is resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011]. NEED A LAST SENTENCE TO WRAP IT UP, HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY. PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP PREDATOR (I.E. NO CONSUMERS...LOOK AT OLD TEXT FOR WORDING SUGGESTIONS.)

@@

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: *CONDENSE THIS. NO NEED TO GIVE ALL THE DETAILS YET. RIGHT NOW IT IS JUST CONFUSING.* (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972; @Lawlor1976); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986; McPeekRECENT]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

@@

To examine how ecological character displacement affects resource abundances and food-web stability<!--consumer-resource interactions and food-web dynamics-->, I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the simplest scenario, consumers can forage for both resources simultaneously and their feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$ [@MacArthur1972]. The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail [@Lawlor1976; @Abrams1986], with the general result being divergent character displacement. I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, measured as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer. I focus here on the ecological consequences of this adaptive divergence.

@@

I examine the effects of character displacement on food-web dynamics by (i) deriving analytical expressions for the relationship between consumer attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement. <!--, rather than any dependence on inherent asymmetries between consumers or resources.-->

@@

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:MacArthur_Plot}A) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:MacArthur_Plot}A). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:MacArthur_Plot}A). Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0).

@@

```{r Tradeoff, include=FALSE, fig.pos="H", fig.cap="\\label{fig:Tradeoff}Tradeoff forms for consumer attack rates."}
initial_points <- bind_rows(params_2C_2R_MacArthur_linear, params_2C_2R_MacArthur_concaveDN, params_2C_2R_MacArthur_concaveUP) %>% mutate(n = c(n_linear, n_concaveDN, n_concaveUP))

tradeoff_plot <- bind_rows(concaveDN_df, linear_df, concaveUP_df) %>%
  ggplot(., aes(x=aii, y=aij, color = factor(n))) + 
  geom_line() + 
  geom_point(data = initial_points, aes(x=a11, y=a12)) +
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)

@@

My simulations show that the shape of the attack-rate tradeoff qualitatively affects the relationship between character displacement and resource density (Fig. \ref{fig:MacArthur_Plot}). For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:MacArthur_Plot}B) and character displacement has no effect on resource densities (Fig. \ref{fig:MacArthur_Plot}C). If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:MacArthur_Plot}C). This is because the total attack rate of consumers is maximized at intermediate values ($a_{ii}=a_{ij}$) and decreases as consumers diverge (Fig. \ref{fig:MacArthur_Plot}B). When the tradeoff is concave up (orange lines), character displacement suppresses resource densities due to the increase in total attack rates (Fig. \ref{fig:MacArthur_Plot}B,C). Although the equation I derived for resource density was for the scenario where both consumers and both resources are present, it accurately predicts the density of resources when a single consumer reaches its ESS (triangles on respective colored lines in Fig. \ref{fig:MacArthur_Plot}C). This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:MacArthur_Plot}B), resulting in equivalent resource densities.

The effect of character displacement on resource density closely matches its impact on food-web stability (Fig. \ref{fig:MacArthur_Plot}D). For example, when character displacement results in resource suppression (orange), there is a corresponding decrease in stability. Similarly, if character displacement does not influence resource densities (blue), then there is no corresponding effect on stability. Curiously, this match between impact on resources and stability breakdowns with the concave down tradeoff (green). <!--This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:Stability}). The effect of this eco-evolutionary feedback can be seen by comparing initial (small points) with evolved strategies (large points). For example, when the tradeoff is concave up (orange points), the three-species system (small triangle) is initially more stable than the four-species system (small circle), but this relationship reverses by the time consumers reach their evolutionary stable strategy.-->

```{r MacArthur_Resources, include=FALSE, fig.cap="\\label{fig:MacArthur_Resources}Effect of character displacement on total attack rates (A) and resource densities (B) when consumers can forage for both resources simultaneously."}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=factor(n), shape=Competitor), size=3, color="black") + #, show.legend=FALSE) + size=5,
  #geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=a11+a12, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  ylab(expression(a[ii]+a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability, include=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="black", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=5, color="black", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="black") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r MacArthur Model, include=FALSE}
plot_macarthur_model <- plot_grid(
  tradeoff_plot + theme(legend.position = "none"), 
  Mac_attack  + theme(legend.position = "none"), 
  Mac_resources  + theme(legend.position = "none"),
  Stability_Mac  + theme(legend.position = "none"), ncol=2, align='hv', labels = "AUTO")

@@

```{r MacArthur_Plot, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:MacArthur_Plot}Effect of character displacement when consumers can forage for both resources simultaneously. Panel (A) shows the different tradeoffs in consumer attack rates used in the simulation (green, *n* = 1.15; blue, *n* = 1; orange, *n* = 0.85). Panel (B) shows how the shape of this tradeoff affects the magnitude of character displacement and the evolution of total attack rates in the consumer community. Large circles (two consumers) and triangles (one consumer) correspond to the end points of the eco-evolutionary simulation. Panels (C) and (D) show how different tradeoffs indirectly affect total resource densities (C) and food-web stability (D). Our simulations used the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2. We set an initial value of *a~ii~* = 1.2 and *a~ij~* depended on the value of *n*. We set initial consumer and resource abundances as: *R~i~* = *R~j~* = 2; *C~i~* = *C~j~* = 1."}
knitr::include_graphics("Figure_ECD_MacArthur.pdf")
```

@@

The previous model, although the starting point for many models of competition, does not reflect many food webs where consumers are more mobile than their resources. In this scenario, consumer foraging behavior can link resources that occur in distinct habitats. The only character displacement model that I am aware of that modeled resources in different habitats was one examined by @Lawlor1976. This model takes the same form as the previous model, except now the consumer's feeding rate takes the form:

@@

where $w_{ji}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $j$ is found (i.e. habitat preference). Note that since $w_{ji}$ is a proportion that $w_{jj}=1-w_{ji}$. As with attack rates, I assume that consumer habitat preferences are mirror images of each other ($w=w_{11}=w_{22}$). In @Lawlor1976's detailed evolutionary analysis of this model, they always observed divergent character displacement.

@@

To examine the ecological consequences of this divergence, I again derive the equilibrium solution when both consumers and both resources are present. As before, resource densities are equivalent at this equilibrium ($\hat R = \hat R_1 = \hat R_2$ when both consumers and resources present), but are now determined by the following equation (derivation in supplementary material):

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:LS_Plot}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Plot}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Plot}B). In contrast, resource densities reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Plot}B), because character displacement tends to reach a constraint of complete specialization. It is worth noting that in this foraging scenario, resource densities are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Plot}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Plot}B). <!-- I will revisit this subtle point when I examine the effects of character displacement on food-web stability. -->

As we saw previously, the effect of character displacement on resource densities qualitatively matches its negative effect on food-web stability (Fig. \ref{fig:LS_Plot}C). This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:LS_Plot}C). For example, when the tradeoff is concave up (orange), then the four-species (small circle) food web actually starts off as less stable than the three-species one (small triangle); however, the eco-evolutionary dynamics cause a switch in this pattern once they reach their evolutionary stable strategies (large points).
